---
title: "SpaceMix analysis"
output: html_document
---

## Set up environment
Loading `reticulate` for using python (scikit-allel). To run yourself, you need python v3+ and scikit-allel
```{r, setup}
library(reticulate)
library(tidyverse)
library(here)
library(SpaceMix)
library(viridis)

use_condaenv(condaenv = "genetics", 
             conda = "/home/rstudio/miniconda3/condabin/conda")
```



Get path for vcf file and read in population mapping file. I'm mapping the actual population names to the correct localities.
```{r}
vcf_path <- here("data", "lg_one_per_locus.recode.vcf")

pops <- read_csv(here("data", "pops_file.csv")) %>% 
  mutate(
    site_name = case_when(
      near(longitude, -105.0444, tol = 0.0001) ~ "ST",
      # the near() function is more forgiving than == when comparing doubles
      near(longitude, -105.2503, tol = 0.0001) ~ "PC",
      near(longitude, -105.0980, tol = 0.0001) ~ "LG",
      near(longitude, -104.8789, tol = 0.0001) ~ "P-35",
      near(longitude, -104.8027, tol = 0.0001) ~ "PZ",
      near(longitude, -104.5181, tol = 0.0001) ~ "CH",
      near(longitude, -104.7663, tol = 0.0001) ~ "PT",
      near(longitude, -104.9325, tol = 0.0001) ~ "LS",
      near(longitude, -104.8500, tol = 0.0001) ~ "P-30"
    )
  )
```

Python version being used.
```{python}
import platform
print(platform.python_version())
```

Read in the vcf and take a look at  the keys present in the file.
```{python}
import allel
import numpy as np
import pandas as pd
vcf = allel.read_vcf(r.vcf_path) 
sorted(vcf.keys())
```

## Allele counts and sample size


### Process vcf
Let's take a look at some of the sample names and assign them to a vector for population mapping.
```{python}
samples = vcf["samples"]
vcf["samples"][1:5,]
```

I'm checking that the order of individuals in the `pops` file match the order in the vcf. It does! I can safely map the pops to the individuals for allele counting.
```{r}
sum(py$samples == pops$individuals)
```



Extract the genotype array and take a look at it. 
```{python}
gt = allel.GenotypeArray(vcf['calldata/GT'])

gt
```

Filter genotype array for biallelic loci.
```{python}
ac = gt.count_alleles()
ac_bi = ac.is_biallelic()
gt = gt[ac_bi,]
gt
```


Get indices for each site name.
```{r}
# have to subtract by one since these are going to be used as indices in python

st <- which(pops$site_name == "ST") - 1 %>% 
  as.integer()
pc <- which(pops$site_name == "PC") - 1 %>% 
  as.integer()
lg <- which(pops$site_name == "LG") - 1 %>% 
  as.integer()
p_35 <- which(pops$site_name == "P-35") - 1 %>% 
  as.integer()
pz <- which(pops$site_name == "PZ") - 1 %>% 
  as.integer()
ch <- which(pops$site_name == "CH") - 1 %>% 
  as.integer()
pt <- which(pops$site_name == "PT") - 1 %>% 
  as.integer()
ls <- which(pops$site_name == "LS") - 1 %>% 
  as.integer()
p_30 <- which(pops$site_name == "P-30") - 1 %>% 
  as.integer()


```

Create a dictionary to map populations to their indices.
```{python}
pops_dict = {"ST" : r.st, "PC": r.pc, "LG": r.lg, "P_35": r.p_35, "PZ": r.pz, "CH": r.ch, "PT": r.pt, "LS": r.ls, "P_30": r.p_30}
pops_dict
```

### Allele counts
Get an allele counts table. 
```{python}
ac_pops = gt.count_alleles_subpops(subpops = pops_dict, max_allele=1)

# We're only interested in the alternative allele counts for particular loci
alt_pops = {}
for key in ac_pops:
   alt_pops[key] = ac_pops[key][:,1]

```


Convert allele counts to a matrix. 
```{r}
ac_mat <- bind_rows(py$alt_pops) %>% 
  as.matrix() %>% 
  t()

dim(ac_mat)
ac_mat[1:5, 1:5]
```



### Sample size
Get sample size matrix. For each locus within each subpopulation I am summing the reference and alternate allele counts. This only sums the alleles that were genotyped (i.e. doesn't include missing data).

```{python}
sample_size = {}
for key in ac_pops:
  sample_size[key] = np.empty(shape = (len(ac_pops[key]), 1))
  for loc in range(len(ac_pops[key])):
    sample_size[key][loc,] = sum(ac_pops[key][loc,])

print("Sample size for a single pop (ST)")
sample_size["ST"]
```


Convert sample sizes dictionary to a matrix
```{r}
ss_mat <- bind_rows(py$sample_size) %>% 
  as.matrix() %>% 
  t()

dim(ss_mat)
ss_mat[1:5, 1:5]
```


### Coordinates
Convert the coordinates into a two-column matrix.
```{r}
pops_unique <- filter(pops, !duplicated(site_name))
long <- pops_unique$longitude
lat <- pops_unique$latitude

locs <- cbind(long, lat)

locs
```

## Run SpaceMix

Note- can't attempt to specify another directory to save files to because `run.spacemix.analysis()` tries to `setwd()`, which R Projects don't like.

### Runs 1 and 2: No migration, spatial prior
This run is inferring the geogenetic locations of individuals without admixture. I'm using observed locations as spatial priors. Repeated this twice.  

#### Run 1
```{r}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "target",
  long.model.option = "target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs[,1],
  spatial.prior.Y.coordinates = locs[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "no_mig_spatial1"
)
```

#### Run 2
```{r}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "target",
  long.model.option = "target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs[,1],
  spatial.prior.Y.coordinates = locs[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "no_mig_spatial2"
)
```

### Run 3: No migration, random prior

Randomly sample coordinates from a uniform distribution
```{r}
set.seed(42427)
long_rand <- runif(9, 
                   min = min(locs[,1]), 
                   max = max(locs[,1])) %>% round(4)

lat_rand <- runif(9, 
                   min = min(locs[,2]), 
                   max = max(locs[,2])) %>% round(4)

locs_rand <- cbind(long_rand, lat_rand)
```


```{r}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "target",
  long.model.option = "target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs_rand[,1],
  spatial.prior.Y.coordinates = locs_rand[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "no_mig_rand1"
)
```

### Runs 4 and 5: Migration, spatial prior

#### Run 4
```{r}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "source_and_target",
  long.model.option = "source_and_target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs[,1],
  spatial.prior.Y.coordinates = locs[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "mig_spatial1"
)
```

#### Run 5
```{r}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "source_and_target",
  long.model.option = "source_and_target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs[,1],
  spatial.prior.Y.coordinates = locs[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "mig_spatial2"
)
```

### Run 6: Migration, random prior

Randomly sample coordinates from a uniform distribution. Using a new seed so the coordinates are different from the no migration run.

```{r}
set.seed(327)
long_rand_2 <- runif(9, 
                   min = min(locs[,1]), 
                   max = max(locs[,1])) %>% round(4)

lat_rand_2 <- runif(9, 
                   min = min(locs[,2]), 
                   max = max(locs[,2])) %>% round(4)

locs_rand_2 <- cbind(long_rand_2, lat_rand_2)
```

```{r, echo = FALSE}
run.spacemix.analysis(
  n.fast.reps = 5,
  fast.MCMC.ngen = 5e6,
  fast.model.option = "source_and_target",
  long.model.option = "source_and_target",
  data.type = "counts",
  sample.frequencies = NULL,
  mean.sample.sizes = NULL,
  counts = ac_mat,
  sample.sizes = ss_mat,
  sample.covariance = NULL,
  target.spatial.prior.scale = NULL,
  source.spatial.prior.scale = NULL,
  spatial.prior.X.coordinates = locs_rand_2[,1],
  spatial.prior.Y.coordinates = locs_rand_2[,2],
  round.earth = TRUE,
  long.run.initial.parameters = NULL,
  k = nrow(ss_mat),
  loci = ncol(ss_mat),
  ngen = 1e8,
  printfreq = 1e4,
  samplefreq = 1e5,
  mixing.diagn.freq = 50,
  savefreq = 1e6,
  directory = NULL,
  prefix = "mig_rand1"
)
```



## Evaluate MCMC performance

Read in model object
```{r}
load(here("output", "trial_run1", "trial_run1_LongRun", "trial_run1_space_MCMC_output1.Robj"))
```

Visualize posterior probability trace plot.
```{r}
# Trace plot of posterior probability over the MCMC
plot(Prob,xlab="MCMC iterations",ylab="value",
    main="Posterior probability trace plot",type='l')
```

Visualize alpha parameters of the spatial covariance function
```{r}
matplot(t(nugget),type='l',
            xlab="MCMC iterations",ylab="Parameter value",
            main="Trace plot of nugget parameters")
```

Joint marginal plot of `a1` vs `a0`.
```{r}
# Joint marginal plot of a0 and a1
#   colored by where in the MCMC these 
#   parameters took their values
plot(a0,a1,xlab="a0",ylab="a1",
    main="Joint marginal of a0 and a1",pch=20,
    col=adjustcolor(rainbow(1000,start=4/6,end=6/6),0.3))
legend(x="bottomright",pch=19,cex=0.5,
        col=rainbow(1000,start=4/6,end=6/6)[c(1,500,1000)],
        legend=c("Sampled MCMC iteration 1",
                 "Sampled MCMC iteration 500",
                 "Sampled MCMC iteration 1000"))
```

Acceptance rate of a0.
```{r}
# Acceptance rate of a0 over the course of the 
#   MCMC analysis
plot(accept_rates$a0_accept_rate,
        xlab="MCMC iterations",ylab="Acceptance rate",
        main="Acceptance rate of a0",type='l',
        ylim=c(0.35,0.6))
    abline(h=0.44,col="gray",lty=2)
```

Acceptance rate of nuggets
```{r}
matplot(t(accept_rates$nugget_accept_rate),
            xlab="MCMC iterations",ylab="Acceptance rate",
            main="Acceptance rates of nuggets",type='l',
            ylim=c(0.3,0.7))
    abline(h=0.44,col="gray",lty=2)
```

## Evaluate model adequacy

Load the data
```{r}
# first, load the standardized (mean-centered and normalized)
#   allele frequency data object.  This object, which is the 
#   "MCN.frequencies.list" (Mean Centered and Normalized) is 
#   saved in the Long Run directory, and is generated if the 
#   user has specified either allele count or allele frequeny 
#   data. 
#   Note that it is not generated if the user has specified the 
#   sample covariance.

load(here("output", "trial_run1", "trial_run1_LongRun", "trial_run1_MCN.frequencies.list.Robj"))
```


```{r}
# Calculate the sample covariance from the mean centered 
#   and normalized sample allele frequencies.
sample.covariance <- cov(t(MCN.frequencies.list$mean.centered.normalized.sample.frequencies),
                                use="pairwise.complete.obs")

# Create a matrix that will perform a mean-centering 
#   on the parametric covariance matrix
# Then, mean-center the parametric ovariance matrix.
k <- nrow(MCN.frequencies.list$mean.centered.normalized.sample.frequencies)

MC.matrix <- diag(k) - matrix(1/last.params$inv.mean.sample.sizes  / (sum(1/last.params$inv.mean.sample.sizes)),
                              nrow=k,ncol=k,byrow=TRUE)

MC.parametric.covariance <- (MC.matrix) %*% last.params$admixed.covariance %*% 
  t(MC.matrix)

# Compare the sample covariance to the parametric
#   covariance.  Ideally, there will be a very tight correspondence 
#   between the data and the model.  If there is not, it may 
#   be an indication either that the MCMC has not converged on 
#   the stationary distribution or that the process that generated 
#   the data is only poorly approximated by SpaceMix's model.

# The sample and parametric covariances can be plotted 
#   against each other (if model fit is good they should 
#   fall on the x=y red line)
index.matrix <- upper.tri(sample.covariance,diag=TRUE)
plot(sample.covariance[index.matrix], 
    MC.parametric.covariance[index.matrix],
    col=adjustcolor("black",0.3),pch=20,
    xlab="Sample covariance",
    ylab="Parametric covariance",
    main="Model adequacy:\n matrix comparison")
    abline(0,1,col="red")
```

```{r}
# Or the patterns of decay of covariance with 
#   geographic distance can be compared between 
#   the data and the model.
plot(last.params$D[1:k,1:k][index.matrix], 
        sample.covariance[index.matrix],
        pch=19,col="black",
        xlab="geogenetic distance",
        ylab="covariance",
        main="Model adequacy:\n IBD patterns")
        points(ex.output$last.params$D[1:k,1:k][index.matrix], 
                MC.parametric.covariance[index.matrix],col="red",pch=20)
        legend(x="topright",pch=19,col=c(1,2),
                legend=c("observed","model estimate"))
```

## Geogenetic map

Load data for mapping.
```{r}
load(here("output", "trial_run1", "trial_run1_LongRun", "trial_run1_spacemix.data.Robj"))
```


```{r}
sm_map_list <- make.spacemix.map.list(
  MCMC.output.file = here("output", "trial_run1", "trial_run1_LongRun", "trial_run1_space_MCMC_output1.Robj"),
  geographic.locations = locs,
  name.vector = rownames(ac_mat),
  color.vector = viridis(n = nrow(ac_mat)),
  quantile = 0.95,
  burnin = 0
)
```

```{r}
make.spacemix.map(
  spacemix.map.list = sm_map_list,
  text = TRUE,
  ellipses = TRUE,
  source.option = FALSE
)
```

Original map
```{r}
sm_map_list_og <- make.spacemix.map.list(
  MCMC.output.file = here("output", "trial_run1", "trial_run1_LongRun", "trial_run1_space_MCMC_output1.Robj"),
  geographic.locations = locs,
  name.vector = rownames(ac_mat),
  color.vector = viridis(n = nrow(ac_mat)),
  quantile = 0.95,
  burnin = 0
)
make.spacemix.map(
  spacemix.map.list = sm_map_list_og,
  text = TRUE,
  ellipses = FALSE,
  source.option = FALSE
)
```

